---
title: "Weight Lifting Exercise"
author: "Jurijs Jefimovs"
date: "Sunday, October 19, 2014"
output: html_document
---

###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set.

###Executive summary
In this paper was compared three classification algorithms against "Weight Lifting Exercise" dataset. These are:
- Recursive Partitioning (rpart)
- Random Forest (rf)
- k-Nearest Neighbour (knn)

As was expected, the Random Forest has the highest accuracy among these three algorithms with accuracy rate 95%. So that, in sample error rate is 0.05 or 5%, hence, we should expected higher out of sample expected rate. But after project submission I've received 19 of 20 correct answers.

*Notes: I've commented out some execution code, in order to build html faster. 

```{r load_data}
library(caret)

url_train <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_test <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train_file <- "train.csv"
test_file <- "test.csv"
download.file(url_train, destfile = train_file)
download.file(url_test, destfile = test_file)
train_data <- read.csv(train_file, na.strings=c("#DIV/0!", "NA"))
test_data <- read.csv(test_file, na.strings=c("#DIV/0!", "NA"))
set.seed(123)

```


###Exploratory data analysis
```{r explore}
str(test_data)

```
It seems that first five columns (X,user_name and timestamps) can be filtered out. 


###Preprocessing
First step - let's cut out these variables from data.
Next, lets do some data preparation:
Find variables that doesnt give un much information (near zero vars).
Then scale, center, inpute missing values and extract features using principal component analysis.


```{r Preprocess}

train_data <- train_data[,-c(1,2,3,4,5)]
test_data <- test_data[,-c(1,2,3,4,5)]

#checking near zero variables
near_zero_vars <- nearZeroVar(train_data)
train_data <- train_data[,-near_zero_vars]
test_data <- test_data[,-near_zero_vars]

#preprocess
col_types <- sapply(train_data,class)

preObj <- preProcess(train_data[,-119],  method = c("center", "scale", "knnImpute", "pca"), thresh = 0.95, k = 5)

train_data.prep <- predict(preObj, newdata=train_data[,-119])
test_data.prep <- predict(preObj, newdata=test_data[,-119])

train_data.prep$classe <- train_data$classe

```

After test data preparation, lets divide test data into train and validation subsets in proportion 70% and 30% accordingly.

```{r subsets}
inTrain <- createDataPartition(y=train_data.prep$classe, p=0.7, list=FALSE)
train_data.train <- train_data.prep[inTrain,]
train_data.test <- train_data.prep[-inTrain,]

```

Now, we are ready to train the models. I've picked up tree classification methods to compare:
- Recursive Partitioning (rpart)
- Random Forest (rf)
- k-Nearest Neighbour (knn)
After each train, i will validate model againt test set made previously. Then i will compare the models against accuracy of prediction.

### Recursive Partitioning
```{r rpart}
modFit_rpart <- train(classe ~ ., data=train_data.train, method="rpart")
predict_rpart <- predict(modFit_rpart, train_data.test)
confusionMatrix(train_data.test$classe, predict_rpart)

```
In this model accuracy was wery low = 0.3934


### Random Forest
```{r rf}
modFit_rf <- train(classe ~ ., data=train_data.train, method="rf",  proxy = T)
predict_rf <- predict(modFit_rf, train_data.test)
confusionMatrix(train_data.test$classe, predict_rf)
```
Accuracy : 0.9573

### k-Nearest Neighbour
```{r knn}
modFit_knn <- train(classe ~ ., data=train_data.train, method="knn")
predict_knn <- predict(modFit_knn, train_data.test)
confusionMatrix(train_data.test$classe, predict_knn)
```
Accuracy : 0.9286

###Final prediction
After set of predictions and validations i've picked up the model based on accuracy metric. Random Forest model has highest accuracy rate, so it will be used for final prediction.

```{r fin_prediction}
predict.train <- predict(modFit_rf, test_data.prep)

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(predict.train)
```

###Conclusion

